The Decline of US Birth Rates: Analyzing Minimum Wages, Abortion Policies, Education, and Child Support Enforcement

### 2024-02-15 paper/paper.qmd
adjusted reference
graphs fixed

author: 
  - Rayan Awad Alim
  - Maria Mangru
  - Isha Junega
thanks: "Code and data are available at: https://github.com/RayanAlim/Birth_Rate_Analysis"
date: today
date-format: long
abstract: "Melissa S. Kearney, Phillip B. Levine, and Luke Pardue in their paper, 
'The Puzzle of Falling US Birth Rates since the Great Recession', analyze the 
decreasing birth rates within the US. In this paper, we repicate use their data 
and replicate their results of their study. While they focused on examining 
birth rates by demographics, we focus on minimum wages, education, abortion 
poilicy, and child support enforcement. The importance of our research is that 
it highlights the underlying cause of why parents may not to have many children, 
which leads to a decline in birthrates."
format: pdf
number-sections: true
#bibliography: references.bib
editor: 
  markdown: 
    wrap: 72
---

@@

In this paper, we created a linear regression model to determine the
impact of abortion policy on number of birthrates. We analyzed the
affects of education on the number of birthrates by tracking the the
birthrate for each education level separately. Additionally, to gage the
impact of childcare support enforcement on the number of birthrates, we
looked at how the maximum welfare benefits, including Medicaid, affect
low-income individuals. Analyzing the wages, abortion policies,
education, and childcare support important because they give deeper
insights regarding why parents may not prefer to have many children and
therefore may cause a decline in birthrates. The paper is structured to
include the data, results, and the discussion that include our analysis.
Although the code used in the original paper is Strata, we use R to
replicate some of its graphs, as well as to conduct analysis of our own.

@@

## Variables and Summary Statisitcs
With an emphasis on the years 2001 to 2019, we created and improved a collection
of variables which are needed for studying the decrease in birth rates 
throughout the United States. Our results are stronger because we were able to 
record the short-term impacts of various different socioeconomic determinants 
on birth rates during this time. The development of important variables, their 
importance, storage formats, and the interactions between them are described in 
detail below. Graphs, summary statistics, and comments are used to provide 
context.

@@

# Create the bar plot with scaled averages
ggplot(average_births_by_state, aes(x = reorder(stname, -AverageBirths), y = AverageBirths)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme_minimal() +
  labs(title = "Scaled Average Number of Births by State (2001-2019)",
       x = "State",
       y = "Scaled Average Number of Births") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

@@

# Plot the graph with adjusted scales
ggplot(average_data) +
  #geom_line(aes(x = year, y = Average_Minimum_Wage), colour = "blue", size = 1) +
  #geom_line(aes(x = year, y = Average_Births * scaling_factor), colour = "red", size = 1) +
  geom_line(aes(x = year, y = Average_Minimum_Wage), colour = "blue", linewidth = 1) +
  geom_line(aes(x = year, y = Average_Births * scaling_factor), colour = "red", linewidth = 1) +
  scale_y_continuous(
    "Minimum Wage",
    sec.axis = sec_axis(~ . / scaling_factor, name = "Number of Births")
  ) +
  theme_minimal() +
  labs(title = "Comparison of Average Minimum Wage and Birth Rates (2001-2009)",
       x = "Year")

@@

ggplot(data = combined_data_no_delay, aes(x = year, y = relative_births, group = stname, color = stname)) +
  geom_line() +
  labs(title = "Relative Average Number of Births in States Without Delay (2001-2019)",
       x = "Year",
       y = "Relative Average Number of Births per 1000 People",
       color = "State") +
  theme_minimal()
```

@@

ggplot(data = combined_data_with_delay, aes(x = year, y = relative_births, group = stname, color = stname)) +
  geom_line() +
  labs(title = "Relative Average Number of Births in States With Delay (2001-2019)",
       x = "Year",
       y = "Relative Average Number of Births per 1000 People",
       color = "State") +
  theme_minimal()
```

@@

ggplot(data = transition_states_data, aes(x = year, y = relative_births, group = stname, color = stname)) +
  geom_line() +
  geom_vline(data = policy_changes, aes(xintercept = year_implemented, color = stname), linetype="dashed") +
  geom_text(data = policy_changes, aes(x = year_implemented, y = Inf, label = year_implemented), vjust = -0.5, color = "black") +
  scale_x_continuous(breaks = seq(2001, 2019, by = 2)) +
                       labs(title = "Transition to Policy Delay and Relative Average Number of Births (2001-2019)",
                            x = "Year",
                            y = "Relative Average Number of Births per 1000 People",
                            color = "State") +
                       theme_minimal() +
                       theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

@@

# Output the robust summary as a table
kable(coef_df, format = "markdown")

@@

## Abortion Policy
Through the difference-in-difference analysis, the results indicate a
decrease of approximately 0.777 births per 1000 people in North Carolina
post policy implementation relative to TN, which did not implement the
abortion delay policy. This decrease is statistically significant,
suggesting that implementing an abortion delay policy may have a real
effect on birth rates.

@@

A weakness of this paper is a lack of economical context. For example,
we analyze our data independent of events like the Recession that
happened in 2008. Without providing context to current events in the
time period, our paper gives a weaker justification for the decline in
birthrates. Next steps for improving our analysis include not only
providing context for the events happening in the US, but focusing on
the birth rates on a more micro level. Since each state within the US
has its own laws, it will provide more insight to focus on the birth
rates within each state. Additionally, to improve our analysis
furthermore, we can compare the US to other countries that are similar
from an economical standpoint and compare their birth rate to the US.

@@

# References

>>>>

author: 
  - Rayan Awad Alim
  - Maria Mangru
  - Isha Juneja
thanks: "Code and data are available at: https://github.com/RayanAlim/Birth_Rate_Analysis"
date: today
date-format: long
abstract: "Melissa S. Kearney, Phillip B. Levine, and Luke Pardue in their paper, 
'The Puzzle of Falling US Birth Rates since the Great Recession', analyze the 
decreasing birth rates within the US. In this paper, we repicate use their data 
and replicate their results of their study. While they focused on examining 
birth rates by demographics, we focus on minimum wages, education, abortion 
poilicy, and child support enforcement. The importance of our research is that 
it highlights the underlying cause of why parents may not to have many children, 
which leads to a decline in birthrates."
format: pdf
number-sections: true
#bibliography: references.bib
editor: 
  markdown: 
    wrap: 72
---

@@

In this paper, we created a linear regression model to determine the
impact of abortion policy on number of birthrates. We analyzed the
affects of education on the number of birthrates by tracking the the
birthrate for each education level separately. Additionally, to gage the
impact of childcare support enforcement on the number of birthrates, we
looked at how the maximum welfare benefits, including Medicaid, affect
low-income individuals. Analyzing the wages, abortion policies,
education, and childcare support important because they give deeper
insights regarding why parents may not prefer to have many children and
therefore may cause a decline in birthrates. The paper is structured to
include the data, results, and the discussion that include our analysis.
Although the code used in the original paper is Strata, we use R[@citeR]
to replicate some of its graphs, as well as to conduct analysis of our
own.

@@

## Variables and Summary Statisitcs

With an emphasis on the years 2001 to 2019, we created and improved a
collection of variables which are needed for studying the decrease in
birth rates throughout the United States. Our results are stronger
because we were able to record the short-term impacts of various
different socioeconomic determinants on birth rates during this time.
The development of important variables, their importance, storage
formats, and the interactions between them are described in detail
below. Graphs, summary statistics, and comments are used to provide
context.

@@

# Create the bar plot with scaled averages
ggplot(average_births_by_state, aes(x = reorder(stname, -AverageBirths), y = AverageBirths)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme_minimal() +
  labs(title = "Scaled Average Number of Births by State (2001-2019)",
       x = "State",
       y = "Scaled Average Number of Births") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size=7))
```

@@

# Plot the graph with adjusted scales
ggplot(average_data, aes(x = year)) +
  geom_line(aes(y = Average_Minimum_Wage), colour = "blue", linewidth = 1) +
  geom_line(aes(y = Average_Births * scaling_factor), colour = "red", linewidth = 1) +
  scale_y_continuous(
    "Minimum Wage",
    sec.axis = sec_axis(~ . / scaling_factor, name = "Number of Births")
  ) +
  theme_minimal() +
  labs(title = "Comparison of Average Minimum Wage and Birth Rates (2001-2019)",
       x = "Year") +
  theme(plot.title = element_text(size = 10))

@@

ggplot(data = combined_data_no_delay, aes(x = year, y = relative_births, group = stname, color = stname)) +
  geom_line() +
  labs(title = "Relative Average Number of Births in States Without Delay (2001-2019)",
       x = "Year",
       y = "Relative Average Number of Births per 1000 People") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 8, hjust = 0.5), 
    axis.title.y = element_text(size = 8), 
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "left", 
    legend.text = element_text(size = 6), 
    legend.title = element_text(size = 6), 
    legend.margin = margin(t = 5, r = 10, b = 0, l = 10) 
  )

@@

ggplot(data = combined_data_with_delay, aes(x = year, y = relative_births, group = stname, color = stname)) +
  geom_line() +
  labs(title = "Relative Average Number of Births in States With Delay (2001-2019)",
       x = "Year",
       y = "Relative Average Number of Births per 1000 People") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 10),  
    axis.title.y = element_text(size = 10),  
    axis.text.x = element_text(angle = 45, hjust = 1),  
    legend.position = "right",  
    legend.text = element_text(size = 8),  
    legend.title = element_text(size = 10),  
    legend.key.size = unit(0.1, 'cm'),  
    legend.spacing.y = unit(0.1, 'cm')  
  )
```

@@

ggplot(data = transition_states_data, aes(x = year, y = relative_births, group = stname, color = stname)) +
  geom_line() +
  geom_vline(data = policy_changes, aes(xintercept = year_implemented, color = stname), linetype="dashed") +
  geom_text(data = policy_changes, aes(x = year_implemented, y = Inf, label = year_implemented), vjust = -0.5, color = "black") +
  scale_x_continuous(breaks = seq(2001, 2019, by = 2)) +
                       labs(title = "Transition to Policy Delay and Relative Average Number of Births (2001-2019)",
                            x = "Year",
                            y = "Relative Average Number of Births per 1000 People",
                            color = "State") +
                       theme_minimal() +
                       theme(
                         plot.title = element_text(size = 10),
                         axis.title.y = element_text(size = 10),
                         axis.text.x = element_text(angle = 45, hjust = 1)
                         )
```

@@

# Output the robust summary as a table
kable(coef_df, format = "markdown", caption = "Table 1")

@@

## Abortion Policy

Through the difference-in-difference analysis, the results indicate a
decrease of approximately 0.777 births per 1000 people in North Carolina
post policy implementation relative to TN, which did not implement the
abortion delay policy. This decrease is statistically significant,
suggesting that implementing an abortion delay policy may have a real
effect on birth rates.

@@

An improvement we can add to this paper is adding additional economical
context along with our data such as investigating events like the
Recession that took place in 2008, and the conflict regarding
reproductive right policies. Without providing context to current events
in the time period, our paper gives a weaker justification for the
decline in birthrates. Next steps for improving our analysis include not
only providing context for the events happening in the US, but focusing
on the birth rates on a more micro level. Since each state within the US
has its own laws, it will provide more insight to focus on the birth
rates within each state. Additionally, to improve our analysis
furthermore, we can compare the US to other countries that are similar
from an economical standpoint and compare their birth rate to the US.

@@

# References
Grundy E, Murphy M. Demography and public health. In: Detels R, Gulliford M, Karim QA, Tan CC, editors. Oxford textbook of global public health. 6th ed. Oxford, England: Oxford University Press, 718–35. 2015.
